---
title: "Modeling and prediction for movies"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `movies`. Delete this note when before you submit 
your work. 

```{r load-data}
load("movies.Rdata")
```



* * *

## Part 1: Data

Start by exploring the data, and looking at our dependant variable. Our films have a range of different scores, with 50% of scores falling between 5.9 and 7.3 out of 10.

*Get rid of films with a low number of people rating / or missing data

```{r}

str(movies)

summary(movies$imdb_rating)
movies %>% ggplot(aes(x=imdb_rating)) + geom_histogram(binwidth = 1) +ggtitle("Histogram: IMDB Rating")


```



* * *

## Part 2: Research question



* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 4: Modeling

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

*Explain reasons for model selection - probably use a mixture of common sense and then forwards / backwards elimination
*Go through and explain the variables you're going to include in the model


* * *

## Part 5: Prediction

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button above. Make sure that your code is visible in the project you submit. 
Delete this note when before you submit your work.

* * *

## Part 6: Conclusion

